Topological order in the quantum dimer-pentamer model

### 2015-10-14 qdpm_paper.tex
Minor changes and a figure of QDM and QDPM finite size scaling of dimer gap

        The QDPM Hamiltonian comprises the QDM Hamiltonian pluss additional terms coresponding to
        the pentamer dynamics.
        \begin{equation}
            H_{\mathrm{QDPM}} = H_{\mathrm{QDM}} + \mathrm{pentamer\ terms}
        \end{equation}
        Specificaly the penatamers pentamer terms are constructed with the dynamics shown in
        Fig.~\ref{fig:pentamer_moves}.  
        \begin{figure}[htpb]
            \centering
            \includegraphics[width=0.8\linewidth]{pentamer_moves.pdf}
            \caption{a) Horizontal creation an annihilation, b) horizontal move, c) and d) are two
            different possible diagonal moves in the same direction.}
            \label{fig:pentamer_moves}
        \end{figure}
        Each of the  moves shown in Fig.~\ref{fig:pentamer_moves} can be
        rotated to produce the symmetry related counterpart (eg. vertical creation and annihilation). 
        As an example the creation and annihilation term in the Hamiltonian is
        \begin{multline}
        H_{\mathrm{pentamer\ creation}} = t_{\mathrm{create}}
            \sum_{\bstr}
            \left(
                \left|
                    \hprs
                \right\rangle
                \left\langle
                    \hspr
                \right|
                +
                \mathrm{h.c.}
            \right)
            \\
            - v_{\mathrm{create}}
            \sum_{\bstr}
                \left(
                    \left|
                        \hprs
                    \right\rangle
                    \left\langle
                        \hprs
                    \right|
                    +
                    \left|
                        \hspr
                    \right\rangle
                    \left\langle
                        \hspr
                    \right|
                \right)
        \end{multline}
        where the factors $t_{\mathrm{create}}$ and $v_{\mathrm{create}}$ are the coefficients of
        the pentamer pair creation and annihilation respectively. At the RK point
        $t/v=1$ for each all the pentamer terms. 
        Writing the kinetic and potential energy terms as done above for all the dynamics shown in
        Fig.~\ref{fig:pentamer_moves} and all of their symmetry related counterparts comprise the
        the pentamer terms of the Hamiltonian.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Dimer Correlation Results}

    \subsection{Liquid}
    \begin{figure}[htpb]
        \centering
        \includegraphics[width=0.8\linewidth]{dimer_gry_scale_qdpm.pdf}
        \caption{Gray scale (actualy data) of the dimer corelation between the dimer in the upper
        left corner and other dimers on a $16\times16$ lattice.}
        \label{fig:dimer_gry_scale_qdpm}
    \end{figure}

    \subsection{Correlation functions}

    \begin{equation}
        C_{\mathrm{dimer}} = \langle d_0 d_r \rangle - \langle d_0 \rangle   \langle d_r \rangle   
    \end{equation}

    \begin{figure}[htpb]
        \centering
        \includegraphics[width=0.8\linewidth]{spacial_dmr_cor.pdf}
        \caption{(Note: This figure was made using equal probabilities of plaquette flips and
        pentamer moves) Spacial correlation function of parallel dimers for the two different
        sublattices distinguished by their color (red, green).}
        \label{fig:spacial_dmr_cor}
    \end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Estimating the Gap}
    An exponetial decay of the spacial correlations shown in the previous sections 
    suggeste a gap in the spectrum. However we would like to know if the lowest lying excitations are in the vison
    or the dimer sector.  The energy gap of the dimer and vison exsitations can be estimated by investigating the dimer and
    vison imaginary time correlations respectivly as outlined by Henley \cite{Henley2004a}. 
    One changlenge in the gap estimates from the imaginary time correlations is going far enough out
    in imaginary time. In appendix \ref{appendix_gap_err} we discuss our proccess in determining
    that this systematic error is sufficinetly small which is necissary when compairing gap
    estimates for different excitatoins.

@@

    In this paper we only investigate the high symetry points in the Brillouin zone.

    In Table~\ref{tab:gap_table} we show the estimates of the dimer gap, the double $Z_2$ vison gap
    at $L/2$ and the $Z_3$ vison gap.
    \begin{table*}[htpb]
    \setlength\tabcolsep{0.3cm}
        \centering
        \caption{$\Delta \tau = 250$}
        \label{tab:gap_table}
        \begin{tabular}{l l l l}
            \hline\hline
                        & Dimer                 & $Z_2$ double ($L/2$)  &  $Z_3$ single (origin)     \\   
            \hline
            frac 1.0    &  &  & \\
            $8\times8$  & $3.24(5)\times10^{-2}$  &                       &                            \\
            $12\times12$& $1.47(5)\times10^{-2}$ &                       &                            \\
            $16\times16$&                       &                       &                            \\
            $18\times18$&                       &                       &                            \\
            \hline
            frac 0.9    &  &  & \\
            $8\times8$  & $0.3549(9)\times10^{-2}$&$0.0993(3)\times10^{-2}$& $0.3069(3)\times10^{-2}$     \\
            $12\times12$& $0.357(2)\times10^{-2} $&$0.256(2)\times10^{-2} $ & $0.3164(7)\times10^{-2}$    \\
            $16\times16$& $0.357(3)\times10^{-2} $&$0.363(5)\times10^{-2} $ & $0.3170(9)\times10^{-2}$    \\
            $18\times18$& $0.352(2)\times10^{-2} $&$0.419(3)\times10^{-2} $ & $0.3167(7)\times10^{-2}$    \\
            \hline
            frac 0.5    &  &  & \\
            $8\times8$  & $1.0(6)\times10^{-2}$ &$0.05(5)\times10^{-2}$ & $1.7(1)\times10^{-2}$     \\
            $12\times12$& $1.5(7)\times10^{-2}$ &                       & $1.9(7)\times10^{-2}$      \\
            $16\times16$&                       &                       &                            \\
            $18\times18$&                       &                       &                            \\
        \end{tabular}
    \end{table*}

    \subsection{Imaginary time vison correlations}
    \begin{figure}[htpb]
        \centering
        \includegraphics[width=0.8\linewidth]{z2_origin_vison_time_cor.pdf}
        \caption{Imaginary time $z_2$ \textit{single} vison correlations at the origin in the QDM model (red) and QDPM
        model (green) on an $8\times8$ lattice. For the QDPM the dimer/pentamer move fraction is $0.9$.}
        \label{fig:name}
    \end{figure}

    \subsection{Imaginary time dimer correlations}
    \begin{figure}[htpb]
        \centering
        \includegraphics[width=0.8\linewidth]{dimer_origin_time_cor.pdf}
        \caption{Imaginary time dimer correlations of a single dimer in the QDM model (red) and QDPM
        model (green) on an $8\times8$ lattice. For the QDPM the dimer/pentamer move fraction is $0.9$. }
        \label{fig:}

    \end{figure}

>>>>

        The QDPM Hamiltonian comprises the QDM Hamiltonian pluss additional terms coresponding to
        the pentamer dynamics.
        \begin{equation}
            H_{\mathrm{QDPM}} = H_{\mathrm{QDM}} + (1-\gamma)\mathrm{pentamer\ terms}
            ,
        \end{equation}
        where $\gamma$ is a single parameter that tunes the relative energy scale of the petamer
        terms to the QDM terms in the Hamiltonian.

        The penatamers pentamer terms are constructed with the dynamics shown in
        Fig.~\ref{fig:pentamer_moves}.  
        \begin{figure}[htpb]
            \centering
            \includegraphics[width=0.8\linewidth]{pentamer_moves.pdf}
            \caption{a) Horizontal creation an annihilation, b) horizontal move, c) and d) are two
            different possible diagonal moves in the same direction.}
            \label{fig:pentamer_moves}
        \end{figure}
        Each of the  moves shown in Fig.~\ref{fig:pentamer_moves} can be
        rotated to produce the symmetry related counterpart (eg. vertical creation and annihilation). 
        As an example the creation and annihilation term in the Hamiltonian is
        \begin{multline}
        H_{\mathrm{pentamer\ creation}} = t_{\mathrm{create}}
            \sum_{\bstr}
            \left(
                \left|
                    \hprs
                \right\rangle
                \left\langle
                    \hspr
                \right|
                +
                \mathrm{h.c.}
            \right)
            \\
            - v_{\mathrm{create}}
            \sum_{\bstr}
                \left(
                    \left|
                        \hprs
                    \right\rangle
                    \left\langle
                        \hprs
                    \right|
                    +
                    \left|
                        \hspr
                    \right\rangle
                    \left\langle
                        \hspr
                    \right|
                \right)
        \end{multline}
        where the factors $t_{\mathrm{create}}$ and $v_{\mathrm{create}}$ are the coefficients of
        the pentamer pair creation and annihilation respectively. At the RK point
        $t/v=1$ for each all the pentamer terms. 
        Writing the kinetic and potential energy terms as done above for all the dynamics shown in
        Fig.~\ref{fig:pentamer_moves} and all of their symmetry related counterparts comprise the
        the pentamer terms of the Hamiltonian.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Spacial Dimer Correlation Results}

    In this section we discuss the dimer dimer correlation function
    \begin{equation}
        C_{\mathrm{dimer}} = \langle d_0 d_r \rangle - \langle d_0 \rangle   \langle d_r \rangle   
        ,
    \end{equation}
    where $d_i$ is the ising degrees of freedom on link i.
    In order for the QDPM to exhibit topological order there we must find that it is in a gaped
    dimer liquid phase. It is likely that the QDPM in a liquid phase at the RK point because there 
    we find no evedence of broken symmetries in the system sizes up to $32x32$. As an example of the
    absence of broken symmetries we show
    the dimer correlation function measuered between a single dimer and all other dimers on the
    lattice in Fig.~\ref{fig:fig:dimer_gry_scale_qdpm} for a $16x16$ lattice.
    \subsection{Liquid}
    \begin{figure}[htpb]
        \centering
        \includegraphics[width=0.8\linewidth]{dimer_gry_scale_qdpm.pdf}
        \caption{Gray scale (actualy data) of the dimer corelation between the dimer in the upper
        left corner (colored red) and other dimers on a $16\times16$ lattice.}
        \label{fig:dimer_gry_scale_qdpm}
    \end{figure}

    We also find exponetialy decaying dimer correlations 
    In Fig.~\ref{fig:fig:spacial_dmr_cor} we show the dimer correlations between horizontal dimers
    in a vertical column. We find the correlation length to be $HERE$.
    \begin{figure}[htpb]
        \centering
        \includegraphics[width=0.8\linewidth]{spacial_dmr_cor.pdf}
        \caption{(Note: This figure was made using equal probabilities of plaquette flips and
        pentamer moves) Spacial correlation function of parallel dimers for the two different
        sublattices distinguished by their color (red, green).}
        \label{fig:spacial_dmr_cor}
    \end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Estimating the Gap}
    The exponetial decay of the spacial correlations shown in the previous sections 
    suggest a gap in the spectrum though it is important to determine if the lowest lying
    excitations are in the vison-like
    or the dimer-like.  The energy gap of the dimer-like and vison-like excitations can be estimated 
    by investigating the dimer and
    vison imaginary time correlations respectivly as outlined by Henley \cite{Henley2004a}. 
    One changlenge in this type of gap estimate is in resloving the correlation functions far enough out
    in imaginary time. In appendix \ref{appendix_gap_err} we discuss our methods used to determine
    that this particular source of systematic error is sufficinetly comparied to the gap
    estimates for different types of excitatoins.

@@

    In this paper we only investigate the gap at high symetry points in the Brillouin zone. 
    In Table~\ref{tab:gap_table} we show the estimates of the dimer gap, the double $Z_2$ vison gap
    at $L/2$ and the $Z_3$ vison gap.
    \begin{table*}[htpb]
    \setlength\tabcolsep{0.3cm}
        \centering
        %\caption{$\Delta \tau = ??$}
        \label{tab:gap_table}
        \begin{tabular}{l l l l}
            \hline\hline
                        & Dimer                 & $Z_2$ double ($L/2$)  &  $Z_3$ single (origin)     \\   
            \hline
            $\gamma=1.0$ $(\Delta \tau =250)$    &  &  & \\
            $8\times8$  & $3.24(5)\times10^{-2}$ &                       &                            \\
            $12\times12$& $1.47(5)\times10^{-2}$ &                       &                            \\
            $16\times16$& $0.66(3)\times10^{-2}$ &                       &                            \\
            $18\times18$& $0.64(2)\times10^{-2}$ &                       &                            \\
            \hline
            $\gamma=0.9$ $(\Delta \tau =500)$    &  &  & \\
            $8\times8$  & $0.3549(9)\times10^{-2}$&$0.0993(3)\times10^{-2}$& $0.3069(3)\times10^{-2}$     \\
            $12\times12$& $0.357(2)\times10^{-2} $&$0.256(2)\times10^{-2} $ & $0.3164(7)\times10^{-2}$    \\
            $16\times16$& $0.357(3)\times10^{-2} $&$0.363(5)\times10^{-2} $ & $0.3170(9)\times10^{-2}$    \\
            $18\times18$& $0.352(2)\times10^{-2} $&$0.419(3)\times10^{-2} $ & $0.3167(7)\times10^{-2}$    \\
            \hline
            $\gamma=0.5$    & & & \\
            $8\times8$  & $1.0(6)\times10^{-2}$ &$0.05(5)\times10^{-2}$ & $1.7(1)\times10^{-2}$     \\
            $12\times12$& $1.5(7)\times10^{-2}$ &                       & $1.9(7)\times10^{-2}$      \\
            $16\times16$&                       &                       &                            \\
            $18\times18$&                       &                       &                            \\
        \end{tabular}
    \end{table*}
    This table shows that the QDPM gap is relativly constant as the system size is increased and
    that the vison gap is lower than the dimer gap. In Fig.~\ref{fig:qdm_qdpm_gap} we show the
    finite size scaling of the dimer gap for both the QDM and the QDPM.
    \begin{figure}[htpb]
        \centering
        \includegraphics[width=0.8\linewidth]{qdm_vs_qdpm_dmr_gap.pdf}
        \caption{Here we show the finite size scaling of the dimer gap for both the QDM and the
        QDPM.}
        \label{fig:qdm_qdpm_gap}
    \end{figure}

%    \subsection{Imaginary time vison correlations}
%    \begin{figure}[htpb]
%        \centering
%        \includegraphics[width=0.8\linewidth]{z2_origin_vison_time_cor.pdf}
%        \caption{Imaginary time $z_2$ \textit{single} vison correlations at the origin in the QDM model (red) and QDPM
%        model (green) on an $8\times8$ lattice. For the QDPM the dimer/pentamer move fraction is $0.9$.}
%        \label{fig:name}
%    \end{figure}
%
%    \subsection{Imaginary time dimer correlations}
%    \begin{figure}[htpb]
%        \centering
%        \includegraphics[width=0.8\linewidth]{dimer_origin_time_cor.pdf}
%        \caption{Imaginary time dimer correlations of a single dimer in the QDM model (red) and QDPM
%        model (green) on an $8\times8$ lattice. For the QDPM the dimer/pentamer move fraction is $0.9$. }
%        \label{fig:}
%
%    \end{figure}
